# Advanced Analysis Modules {#advanced-analysis}

This chapter covers the advanced statistical analysis modules in **UKBAnalytica**, including subgroup analysis, propensity score methods, mediation analysis, and multiple imputation pooling.

## Overview

| Module | File | Purpose |
|--------|------|---------|
| Subgroup Analysis | `subgroup.R` | Stratified analysis with interaction p-values |
| Propensity Score | `propensity.R` | PSM matching and IPTW weighting |
| Mediation Analysis | `mediation.R` | Causal mediation (based on regmedint) |
| MI Pooling | `mi_pool.R` | Multiple imputation result combining |
| Visualization | `visualization.R` | Forest plots, K-M curves, diagnostic plots |

---

## Subgroup Analysis {#subgroup}

Subgroup (stratified) analysis examines whether treatment effects vary across patient subgroups and calculates interaction p-values.

### Basic Usage

```{r subgroup-basic, eval=FALSE}
library(UKBAnalytica)

# Run subgroup analysis by age group
result <- run_subgroup_analysis(
  data = dt,
  exposure = "treatment",
  outcome = "event",
  subgroup_var = "age_group",
  covariates = c("sex", "bmi"),
  model_type = "cox",
  endpoint = c("time", "status")
)

print(result)
#>     subgroup  n n_event    HR lower95 upper95  pvalue p_interaction
#> 1  <50 years 500     45  1.25    0.85    1.84   0.256         0.032
#> 2 50-65 years 800     92  1.58    1.12    2.23   0.009         0.032
#> 3  >65 years 400     78  0.89    0.62    1.28   0.532         0.032
```

### Multiple Subgroup Variables

```{r subgroup-multi, eval=FALSE}
# Analyze multiple subgroup variables at once
results <- run_multi_subgroup(
  data = dt,
  exposure = "treatment",
  outcome = "event",
  subgroup_vars = c("age_group", "sex", "smoking_status"),
  covariates = c("bmi", "diabetes"),
  model_type = "cox",
  endpoint = c("time", "status")
)

# Create forest plot
plot_forest(results, title = "Subgroup Analysis: Treatment Effect")
```

### Parameters

| Parameter | Description |
|-----------|-------------|
| `exposure` | Treatment/exposure variable name |
| `outcome` | Outcome variable (for logistic/linear) |
| `subgroup_var` | Variable defining subgroups |
| `model_type` | `"cox"`, `"logistic"`, or `"linear"` |
| `endpoint` | For Cox: `c("time_var", "status_var")` |
| `ref_level` | Reference level for subgroup variable |

---

## Propensity Score Methods {#propensity}

Propensity score methods help control for confounding in observational studies.

### Estimate Propensity Scores

```{r ps-estimate, eval=FALSE}
# Estimate propensity scores using logistic regression
ps_result <- estimate_propensity_score(
  data = dt,
  treatment = "treatment",
  covariates = c("age", "sex", "bmi", "smoking", "diabetes"),
  method = "logistic"  # or "gbm" for gradient boosting
)

# Add PS to data
dt$ps <- ps_result$ps

# Visualize PS distribution
plot_ps_distribution(ps_result, type = "histogram")
plot_ps_distribution(ps_result, type = "density")
```

### Propensity Score Matching (PSM)

```{r psm, eval=FALSE}
# Perform 1:1 nearest neighbor matching
matched <- match_propensity(
  data = dt,
  treatment = "treatment",
  ps = dt$ps,
  ratio = 1,
  caliper = 0.2,  # 0.2 * SD of PS
  method = "nearest"
)

# Check balance
balance <- assess_balance(
  data = matched$matched_data,
  treatment = "treatment",
  covariates = c("age", "sex", "bmi", "smoking", "diabetes")
)
print(balance)

# Visualize balance improvement
plot_balance(balance, title = "Covariate Balance: Before vs After Matching")
```

### Inverse Probability of Treatment Weighting (IPTW)

```{r iptw, eval=FALSE}
# Calculate IPTW weights
dt$weight <- calculate_weights(
  ps = dt$ps,
  treatment = dt$treatment,
  type = "ate"  # or "att", "atc"
)

# Run weighted analysis
result <- run_weighted_analysis(
  data = dt,
  treatment = "treatment",
  outcome = "event",
  weights = "weight",
  model_type = "cox",
  endpoint = c("time", "status")
)
print(result)
```

### Weight Types

| Type | Formula | Use Case |
|------|---------|----------|
| `"ate"` | T/PS + (1-T)/(1-PS) | Average Treatment Effect |
| `"att"` | T + (1-T)*PS/(1-PS) | Effect on Treated |
| `"atc"` | T*(1-PS)/PS + (1-T) | Effect on Controls |

---

## Mediation Analysis {#mediation}

Causal mediation analysis decomposes the total effect into direct and indirect effects, using the `regmedint` package as backend.

### Basic Mediation Analysis

```{r mediation-basic, eval=FALSE}
# Run mediation analysis
med_result <- run_mediation(
  data = dt,
  exposure = "treatment",
  mediator = "inflammation",
  outcome = "cvd_event",
  covariates = c("age", "sex", "bmi"),
  outcome_type = "cox",        # "linear", "logistic", or "cox"
  mediator_type = "linear",    # "linear" or "logistic"
  event_var = "status",        # For Cox models
  time_var = "time",           # For Cox models
  a0 = 0, a1 = 1,              # Exposure contrast
  m_cde = 0,                   # Mediator value for CDE
  boot = TRUE,                 # Bootstrap CIs
  nboot = 1000
)

# View results
summary(med_result)
#> Mediation Analysis Results
#> ==========================
#> Exposure: treatment (0 vs 1)
#> Mediator: inflammation
#> Outcome: cvd_event (Cox model)
#>
#>         Effect  Estimate   Std.Err   95% CI          P-value
#>         CDE     0.234      0.089     [0.060, 0.408]  0.008
#>         PNDE    0.198      0.082     [0.037, 0.359]  0.016
#>         TNIE    0.156      0.045     [0.068, 0.244]  <0.001
#>         TE      0.354      0.095     [0.168, 0.540]  <0.001
#>         PM      44.1%      -         -               -
```

### Effect Definitions

| Effect | Description |
|--------|-------------|
| CDE | Controlled Direct Effect (setting M=m) |
| PNDE | Pure Natural Direct Effect |
| TNIE | Total Natural Indirect Effect |
| TE | Total Effect (PNDE + TNIE) |
| PM | Proportion Mediated (TNIE/TE) |

### Multiple Mediators

```{r mediation-multi, eval=FALSE}
# Test multiple mediators
multi_results <- run_multi_mediator(
  data = dt,
  exposure = "treatment",
  mediators = c("inflammation", "oxidative_stress", "insulin_resistance"),
  outcome = "cvd_event",
  covariates = c("age", "sex"),
  outcome_type = "cox",
  event_var = "status",
  time_var = "time"
)

# Forest plot of indirect effects
plot_mediation_forest(multi_results, effect_type = "tnie")
```

### Visualization

```{r mediation-plot, eval=FALSE}
# Effect bar chart
plot_mediation(med_result, type = "effects")

# Effect decomposition
plot_mediation(med_result, type = "decomposition")

# Path diagram
plot_mediation(med_result, type = "path")
```

---

## Multiple Imputation Pooling {#mi-pooling}

When using multiple imputation for missing data, results from each imputed dataset must be combined using Rubin's Rules.

### Basic Usage

```{r mi-basic, eval=FALSE}
# Assume mi_datasets is a list of imputed datasets from mice
# mi_data <- mice(original_data, m = 20)
# mi_datasets <- complete(mi_data, action = "all")

# Method 1: Fit models first, then pool
models <- lapply(mi_datasets, function(d) {
  coxph(Surv(time, status) ~ treatment + age + sex, data = d)
})

pooled <- pool_mi_models(models, model_type = "cox")
summary(pooled)
#> Multiple Imputation Results (Rubin's Rules)
#> ============================================
#> Number of imputations: 20
#> Model type: cox
#>
#>   Term          HR     Std.Err   95% CI            p-value   FMI
#>   treatment     1.456  0.156     [1.072, 1.978]    0.019     7.8%
#>   age           1.023  0.008     [1.007, 1.039]    0.004     4.5%
#>   sex           0.856  0.112     [0.687, 1.067]    0.168     5.2%
```

### One-Step Fitting and Pooling

```{r mi-onestep, eval=FALSE}
# Method 2: Fit and pool in one call
pooled <- pool_mi_models(
  datasets = mi_datasets,
  formula = Surv(time, status) ~ treatment + age + sex + bmi,
  model_type = "cox"
)

# Logistic regression example
pooled_logit <- pool_mi_models(
  datasets = mi_datasets,
  formula = diabetes ~ treatment + age + sex + bmi,
  model_type = "logistic",
  exponentiate = TRUE  # Output OR
)
```

### Supported Model Types

| Model | Function | Output |
|-------|----------|--------|
| `"lm"` | `lm()` | Beta coefficients |
| `"logistic"` | `glm(family=binomial)` | Odds Ratio |
| `"poisson"` | `glm(family=poisson)` | Rate Ratio |
| `"cox"` | `coxph()` | Hazard Ratio |
| `"negbin"` | `glm.nb()` | Rate Ratio |

### Custom Estimates Pooling

```{r mi-custom, eval=FALSE}
# Pool any custom estimates (not just regression)
mean_diffs <- lapply(mi_datasets, function(d) {
  mean(d$outcome[d$treatment == 1]) - mean(d$outcome[d$treatment == 0])
})

# Estimate variances
vars <- lapply(mi_datasets, function(d) {
  n1 <- sum(d$treatment == 1)
  n0 <- sum(d$treatment == 0)
  var(d$outcome[d$treatment == 1])/n1 + var(d$outcome[d$treatment == 0])/n0
})

pooled_diff <- pool_custom_estimates(
  estimates = mean_diffs,
  variances = lapply(vars, function(v) matrix(v)),
  labels = "Mean Difference"
)
summary(pooled_diff)
```

### Visualization

```{r mi-viz, eval=FALSE}
# Forest plot of pooled estimates
plot_mi_pooled(pooled, exponentiate = TRUE)

# Diagnostic plot: Fraction of Missing Information
plot_mi_diagnostics(pooled, type = "fmi")

# Variance ratio diagnostic
plot_mi_diagnostics(pooled, type = "variance_ratio")
```

### Understanding FMI

The **Fraction of Missing Information (FMI)** indicates how much the missing data affects each estimate:

| FMI Level | Interpretation |
|-----------|----------------|
| < 10% | Low impact from missing data |
| 10-30% | Moderate impact |
| 30-50% | High impact |
| > 50% | Very high impact; consider more imputations |

---

## Visualization Functions {#visualization}

### Forest Plot

```{r viz-forest, eval=FALSE}
# From subgroup analysis
plot_forest(
  subgroup_results,
  estimate_col = "HR",
  lower_col = "lower95",
  upper_col = "upper95",
  null_value = 1,
  log_scale = TRUE,
  title = "Subgroup Analysis Forest Plot"
)
```

### Kaplan-Meier Curves

```{r viz-km, eval=FALSE}
plot_km_curve(
  data = dt,
  time = "surv_time",
  status = "status",
  group = "treatment",
  title = "Survival by Treatment Group",
  risk_table = TRUE,
  conf_int = TRUE
)
```

### Propensity Score Plots

```{r viz-ps, eval=FALSE}
# PS distribution
plot_ps_distribution(ps_result, type = "density")

# Balance plot
plot_balance(balance_result, title = "SMD Before/After Matching")

# Calibration plot
plot_calibration(predicted = dt$ps, observed = dt$treatment)
```

---

## Integration Example: Complete Workflow {#workflow}

Here's a complete example combining multiple modules:

```{r workflow, eval=FALSE}
library(UKBAnalytica)
library(mice)
library(survival)

# 1. Multiple imputation (using mice)
mi_data <- mice(original_data, m = 20, method = "pmm")
mi_datasets <- complete(mi_data, action = "all")

# 2. Propensity score weighting in each imputation
weighted_models <- lapply(mi_datasets, function(d) {
  # Estimate PS
  ps <- estimate_propensity_score(d, "treatment", 
                                   c("age", "sex", "bmi", "smoking"))$ps
  # Calculate weights
  d$weight <- calculate_weights(ps, d$treatment, "ate")
  
  # Fit weighted Cox model
  coxph(Surv(time, status) ~ treatment, data = d, weights = weight)
})

# 3. Pool results
pooled <- pool_mi_models(weighted_models, model_type = "cox")
summary(pooled)

# 4. Subgroup analysis (on complete-case for simplicity)
subgroup_results <- run_multi_subgroup(
  data = original_data[complete.cases(original_data), ],
  exposure = "treatment",
  outcome = "event",
  subgroup_vars = c("age_group", "sex", "diabetes"),
  model_type = "cox",
  endpoint = c("time", "status")
)

# 5. Visualize
plot_forest(subgroup_results)
plot_mi_pooled(pooled)
```

---

## Summary

This chapter covered four advanced analysis modules:

1. **Subgroup Analysis**: `run_subgroup_analysis()`, `run_multi_subgroup()`
2. **Propensity Score**: `estimate_propensity_score()`, `match_propensity()`, `calculate_weights()`
3. **Mediation Analysis**: `run_mediation()`, `run_multi_mediator()`
4. **MI Pooling**: `pool_mi_models()`, `pool_custom_estimates()`

Each module includes dedicated visualization functions and integrates with the standard UKBAnalytica workflow.

### Required Packages

| Module | Required Package |
|--------|------------------|
| Propensity Score Matching | `MatchIt` |
| GBM for PS | `gbm` |
| Mediation Analysis | `regmedint` |
| MI Pooling | `mitools` |
| Cox Models | `survival` |

Install with:
```{r install-deps, eval=FALSE}
install.packages(c("MatchIt", "gbm", "regmedint", "mitools", "survival"))
```
